Improving Direct-Control Reinforcement Learning for Network Intrusion Prevention (WIP)

### 2018-09-02 scone-talk/scone-talk.tex
Finish talk with maths.

@@

\begin{frame}{Multiagent RL for DDoS prevention}
\begin{columns}
	\begin{column}{0.45\linewidth}
		\begin{itemize}
			\item Algorithm: Semi-gradient Sarsa, linear fn approx.
			\item Actions: Drop [0, 10, ... 90]\si{\percent} upstream traffic.
			\item State: load vectors of agent and parents ($\mathbb{R}^4$) $\rightarrow$ tile-coded (fixed-weight binary vector).
			\item Rewards: \num{-1} if $\mathit{load} > U_s$, else fraction of surviving legit traffic.
		\end{itemize}
	\end{column}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\centering
			\resizebox{\linewidth}{!}{
				\begin{tikzpicture}[
				texts/.style = {text=black},
				labeltexts/.style = {text=uofgsandstone},
				treeline/.style = {draw=uofgburgundy},
				treenode/.style = {texts, circle, centered, fill=white, treeline},
				load/.style = {fill=uofgcobalt},
				external/.style = {fill=uofgrust},
				hideline/.style = {draw=uofgsandstone!40!white},
				hidenode/.style = {treenode, hideline},
				grow'=right
				]
				\node[treenode, label={[texts]above:Server}] (root) {}
				child [treeline] { node [treenode, load, label={[texts]above:Core}] (sswitch) {}
					child [treeline] { node [treenode, load, label={[texts]above:Leader}] (teaml) {} 
						child [treeline] { node [treenode, load, label={[texts]above:Intermediate}] (inter) {}
							child [treeline] { node [treenode, load, label={[texts]above:Agent/Egress}] (agent) {}
								child [treeline] { node [treenode, external] (extern) {}
									child [treeline] { node [treenode, external, label={[texts]above:Host}] (host) {} }
									child [hideline] { node [hidenode] (endhost) {} }
								}
							}
							child [hideline] { node [hidenode] (endagent) {} }
						}
						child [hideline] { node [hidenode] (endinter) {} }
					}
					child [hideline] { node [hidenode] (endteaml) {} }
					edge from parent
					node[below, labeltexts] {$U_s$}
				};

				%\draw[-] (teaml) -- (endteaml);
				\node [labeltexts] (kdots) at ($(teaml)!0.5!(endteaml)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of kdots] {$k$};
				\node [labeltexts] (ldots) at ($(inter)!0.5!(endinter)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of ldots] {$l$};
				\node [labeltexts] (mdots) at ($(agent)!0.5!(endagent)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of mdots] {$m$};
				\node [labeltexts] (ndots) at ($(host)!0.5!(endhost)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of ndots] {$n$};
				\end{tikzpicture}
			}
			\caption{
				Network topology diagram.
				Red nodes are external, blue nodes feature in the state vector.
				Any packet drop occurs when forwarding packets from an egress switch to its parent (intermediate) switch.
			}
		\end{figure}
	\end{column}
\end{columns}
\end{frame}

@@

%			\item More granular $\implies$ should focus on flow stats, \alert{not aggregates} like now! %Alongside this higher-granularity view, reformulation to include flow characteristics in the decision-making process.
		\end{itemize}
	\end{column}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\includegraphics[width=\linewidth]{../plots/online-varyN-binary.pdf}
			\caption{Service quality decreases as actions become less granular (affecting $n$ hosts at once).}
		\end{figure}
	\end{column}
	\end{columns}
\end{frame}

\begin{frame}{The case for finer granularity}
\centering
\includegraphics[width=0.8\linewidth]{../plots/online-varyN-binary.pdf}
\end{frame}

\begin{frame}{On collateral damage}
\begin{columns}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\includegraphics[width=\linewidth]{../plots/online-varyN-nginx.pdf}
			\caption{Explicit UDP traffic matches replayed traces (tcpreplay vs hping3). TCP traffic (nginx) is severely punished.}
		\end{figure}
	\end{column}
	\begin{column}{0.45\linewidth}
		\begin{itemize}
			\item Is the simulation environment of past work complete?
			\item \alert{No.} It's reliant on a numerical simulator, derived from observations of \emph{traces}.
			\item UDP benign traffic similar trend to replayed TCP traces, which matches the original results.
			\item Live TCP responds very badly.
		\end{itemize}
	\end{column}
\end{columns}
\end{frame}

\begin{frame}{On collateral damage}
\centering
\includegraphics[width=0.8\linewidth]{../plots/online-varyN-nginx.pdf}
\end{frame}

@@

\begin{frame}{How can we use these observations? (The Immediate Future)}
\begin{itemize}
	\item \textbf{\alert{Why not take actions on a per-flow basis}}?
	\begin{itemize}
		\item Solves the granularity issues by construction.
		\item Allows different treatment by flow features (i.e., protocol).
	\end{itemize}
	\item Need to rethink state space: more costly computation, but we have room to work in.
	\begin{itemize}
		\item We need any additions to be justified beyond just ``more data'', since \alert{changes affect training time and execution time}.
	\end{itemize}
	\item How do we select flows to act upon?
\end{itemize}
\end{frame}

@@

\item Deriving normal model behaviour from traces.
\begin{itemize}
	\item We only need to simulate specific behaviour to test these enhancements, but that can become more representative.
\end{itemize}
\end{itemize}
\end{frame}

>>>>

\usepackage{bm}

@@

\begin{frame}{Multiagent RL for DDoS prevention}
\begin{columns}
	\begin{column}{0.45\linewidth}
		\begin{itemize}
			\item Algorithm: Semi-gradient Sarsa, linear fn approx, $\epsilon$-greedy selection.
			\item Actions: Drop [0, 10, ... 90]\si{\percent} upstream traffic.
			\item State: load vectors of agent and parents ($\mathbb{R}^4$) $\rightarrow$ tile-coded (fixed-weight binary vector).
			\item Rewards: \num{-1} if $\mathit{load} > U_s$, else fraction of surviving legit traffic.
		\end{itemize}
	\end{column}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\centering
			\resizebox{\linewidth}{!}{
				\begin{tikzpicture}[
				texts/.style = {text=black},
				labeltexts/.style = {text=uofgsandstone},
				treeline/.style = {draw=uofgburgundy},
				treenode/.style = {texts, circle, centered, fill=white, treeline},
				load/.style = {fill=uofgcobalt},
				external/.style = {fill=uofgrust},
				hideline/.style = {draw=uofgsandstone!40!white},
				hidenode/.style = {treenode, hideline},
				grow'=right
				]
				\node[treenode, label={[texts]above:Server}] (root) {}
				child [treeline] { node [treenode, load, label={[texts]above:Core}] (sswitch) {}
					child [treeline] { node [treenode, load, label={[texts]above:Leader}] (teaml) {} 
						child [treeline] { node [treenode, load, label={[texts]above:Intermediate}] (inter) {}
							child [treeline] { node [treenode, load, label={[texts]above:Agent/Egress}] (agent) {}
								child [treeline] { node [treenode, external] (extern) {}
									child [treeline] { node [treenode, external, label={[texts]above:Host}] (host) {} }
									child [hideline] { node [hidenode] (endhost) {} }
								}
							}
							child [hideline] { node [hidenode] (endagent) {} }
						}
						child [hideline] { node [hidenode] (endinter) {} }
					}
					child [hideline] { node [hidenode] (endteaml) {} }
					edge from parent
					node[below, labeltexts] {$U_s$}
				};

				%\draw[-] (teaml) -- (endteaml);
				\node [labeltexts] (kdots) at ($(teaml)!0.5!(endteaml)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of kdots] {$k$};
				\node [labeltexts] (ldots) at ($(inter)!0.5!(endinter)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of ldots] {$l$};
				\node [labeltexts] (mdots) at ($(agent)!0.5!(endagent)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of mdots] {$m$};
				\node [labeltexts] (ndots) at ($(host)!0.5!(endhost)$) {$\rvdots$};
				\node [labeltexts, right = -0.1cm of ndots] {$n$};
				\end{tikzpicture}
			}
			\caption{
				Network topology diagram.
				Red nodes are external, blue nodes feature in the state vector.
				Any packet drop occurs when forwarding packets from an egress switch to its parent (intermediate) switch.
			}
		\end{figure}
	\end{column}
\end{columns}
\end{frame}
\newcommand{\acval}[3]{\ensuremath{\operatorname{\hat{q}}(#1, #2, #3)}}
\newcommand{\wvec}[1]{\ensuremath{\bm{w}_{#1}}}
\begin{frame}{The algorithm?}
% make math easy
\begin{subequations}
A little terse, but the main update rule is:
\begin{gather}
\delta_t = R_{t+1} + \gamma \acval{S_{t+1}}{A_{t+1}}{\wvec{t}} - \acval{S_t}{A_t}{\wvec{t}},\\
\bm{w}_{t+1} = \bm{w}_{t} + \alpha \delta_t \nabla_{\wvec{}}{\acval{S_t}{A_t}{\wvec{t}}}
\end{gather}
with linear function approximation:
\begin{gather}
\acval{s}{a}{\wvec{}} = \wvec{}^{\top} \operatorname{\mathbf{x}}(s, a), \\
\nabla_{\wvec{}}{\acval{s}{a}{\wvec{}}} = \operatorname{\mathbf{x}}(s, a)
\end{gather}
\end{subequations}
\end{frame}

@@

%			\item More granular $\implies$ should focus on flow stats, \alert{not aggregates} like now! %Alongside this higher-granularity view, reformulation to include flow characteristics in the decision-making process.
		\end{itemize}
	\end{column}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\includegraphics[width=\linewidth]{../plots/online-varyN-binary-pres.pdf}
			\caption{Service quality decreases as actions become less granular (affecting $n$ hosts at once).}
		\end{figure}
	\end{column}
	\end{columns}
\end{frame}

\begin{frame}{The case for finer granularity}
\centering
\includegraphics[width=0.8\linewidth]{../plots/online-varyN-binary-pres.pdf}
\end{frame}

\begin{frame}{On collateral damage}
\begin{columns}
	\begin{column}{0.5\linewidth}
		\begin{figure}
			\includegraphics[width=\linewidth]{../plots/online-varyN-nginx-pres.pdf}
			\caption{Explicit UDP traffic matches replayed traces (hping3 vs tcpreplay). TCP traffic (nginx) is severely punished.}
		\end{figure}
	\end{column}
	\begin{column}{0.45\linewidth}
		\begin{itemize}
			\item Is the simulation environment of past work complete?
			\item \alert{No.} It's reliant on a numerical simulator, derived from observations of \emph{traces}.
			\item UDP benign traffic similar trend to replayed TCP traces, which matches the original results.
			\item Live TCP responds very badly.
		\end{itemize}
	\end{column}
\end{columns}
\end{frame}

\begin{frame}{On collateral damage}
\centering
\includegraphics[width=0.8\linewidth]{../plots/online-varyN-nginx-pres.pdf}
\end{frame}

@@

\begin{frame}{How can we use these observations? (The Immediate Future)}
\begin{itemize}
	\item \textbf{\alert{Why not take actions on a per-flow basis?}}
	\begin{itemize}
		\item Solves the granularity issues by construction.
		\item Allows different treatment by flow features (i.e., protocol).
	\end{itemize}
	\item Need to rethink state space: more costly computation, but we have room to work in.
	\begin{itemize}
		\item We need any additions to be justified beyond just ``more data'', since \alert{changes affect training time and execution time}.
	\end{itemize}
	\item How do we select flows to act upon?
\end{itemize}
\end{frame}

@@

\item Deriving normal model behaviour from traces.
\begin{itemize}
	\item We only need to simulate specific behaviour to test these enhancements, but that can become more representative.
\end{itemize}

\item Investigate other RL algorithms.
\begin{itemize}
	\item ``Deep learning'' probably not feasible.
	\item \alert{$\operatorname{TD}(\lambda)$, actor-critic methods}.
\end{itemize}
\end{itemize}
\end{frame}
